---
title: "Creating Interactive Plots in R"
author: "Greg Mitchell and Tara L. Crewe"
date: "September 12, 2016"
output: word_document
---

This code was written by Greg Mitchell and modified into an Rmd file for the Motus website by Tara Crewe.

# Objective and Background

The purpose of this code is to visualize Motus detection data and assess and measure departure times and other parameters. The following function allows the user to zoom in on a plot for data exploration.  Once zoomed in to the data of interest, a secod function allows the user to measure data from the plot, and store the output in a database.  The function defined is for the specific objective of measuring departure time from a site, but it can be modified for other purposes.

The dataset used in this example is from a single tower on Kent Island, New Brunswick, Canada, and includes all detections from 2010 for a bird with tag ID 155. Data were recorded with a Lotek SRX600 receiver.  

```{r setup, message = FALSE}

#setwd("F:/.....") # user may wish to set the working directory
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)

```

```{r import data}

tag_155 <- read.csv("TAG_155.csv")

# modify times and sort data by Date_Time
tag_155$Date_Time <- ymd_hms(tag_155$Date_Time, tz="America/New_York")
tag_155$Date_Time <- with_tz(tag_155$Date_Time, tz="UTC")
tag_155 <- tag_155[order(tag_155$Date_Time),]

```

## Plot the data

Note that for a given tower, you may wish to look at each antenna separately to get a feel for the direction of movement.  In this example there are four antenna, so we split the screen accordingly.


```{r Set plot parameters a}

n.antenna <- length(unique(tag_155$Antenna))
split.screen(c(n.antenna,1))

```
We then set the parameters for each plot, and plot the data.  Here, we are using the same parameters for each panel:

1. 'Mar' refers to the number of lines of text in the margin around the plot on the bottom, left, top, and right, respectively.  
2. 'Cex.axis' refers to text size.  
3. 'mgp' is the margin line on which axes labels are drawn.  
4. 'xpd=T' allows you to plot things like the legend outside of the plot.  
5. 'omi' refers to the size of the outer margins of the plots in inches, and allows for finer control relative to mar.

```{r Plot Data}

screen(1, FALSE)
par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
with(subset(tag_155, Antenna==1), plot(Power ~ Date_Time, type='b', cex=1, col=8, ylab="Power Ant. 1"))

screen(2, FALSE)
par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
with(subset(tag_155, Antenna==2), plot(Power ~ Date_Time, type='b', cex=1, col=2, ylab="Power Ant. 2"))

screen(3, FALSE)
par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
with(subset(tag_155, Antenna==3), plot(Power ~ Date_Time, type='b', cex=1, col=3, ylab="Power Ant. 3"))

screen(4, FALSE)
par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
with(subset(tag_155, Antenna==4), plot(Power ~ Date_Time, type='b', cex=1, col=4, ylab="Power Ant. 4"))

```

# Zooming Function

Now that the data is plotted, we can apply the zooming function. This is a simple function to establish new plotting limits. You will have to adjust the split screen function as necessary. 

```{r zoom function, eval = FALSE}

zoom = function(x) {  #x = name of dataframe
 reg = locator(4)
 reg$x <- reg$x + origin
 yaxis <- c(max(reg$y)+10, min(reg$y)-10)
 xaxis <- c(max(reg$x), min(reg$x)) #Note - I am not sure why, but taking min or max of date changes tz.
 xaxis <- with_tz(xaxis, tz="UTC")
 dev.off()
 
 split.screen(c(n.antenna,1))
 
 screen(1, FALSE)
 par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
 plot(yaxis~xaxis,type='n', ylab="Power Ant. 1")
 with(subset(x, Antenna==1), points(Power ~ Date_Time, type='b', cex=2, col=8))
 
 screen(2, FALSE)
 par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
 plot(yaxis~xaxis,type='n', ylab="Power Ant. 2")
 with(subset(x, Antenna==2), points(Power ~ Date_Time, type='b', cex=2, col=2))
 
 screen(3, FALSE)
 par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
 plot(yaxis~xaxis,type='n', ylab="Power Ant. 3")
 with(subset(x, Antenna==3), points(Power ~ Date_Time, type='b', cex=2, col=3))
 
 screen(4, FALSE)
 par(mar=c(2,4,1,1), cex.axis=1.2, cex.lab=1.2, mgp=c(2.5,1,0), xpd=T, omi=c(0.5,0,0,0), xpd=F)
 plot(yaxis~xaxis,type='n', ylab="Power Ant. 4")
 with(subset(x, Antenna==4), points(Power ~ Date_Time, type='b', cex=2, col=4))
}
```

# Apply Zoom Function

Now we can apply the zoom function to tag 155.  Note by calling zoom(tag_155), we are applying the zoom function to the detection data of tag ID 155.  Before applying the zoom function, you need to point R to the screen (antenna) that you are interested in zooming in on.  Here, we zoom in on antenna 3, and we can only meaningfully interact with a single screen/antenna at a time.  Note that you can keep zooming in further, but each time you wish to zoom, you need to specify both lines of the following code, to specify the screen you want to zoom in on and call the zoom() function.  Otherwise the plots will go blank. In order to zoom, you move the cursor over the panel you've specified and specify the boundaries of the new (zoomed) plot by clicking on the plot four times, in four different areas that encapsulate the data you want to view.  Essentially, you will make a box around the area you want to zoom in on.

```{r zoom in, eval = FALSE}

screen(3, FALSE)
zoom(tag_155)

```

# Measure values from a plot

Once we've sufficiently zoomed in to the plot, we can use the following function to measure departure timing.  This function can be modified to measure other parameters of interest.

```{r measure plot, eval = FALSE}

get_data <- function(x) {
 
 Tag <- readline(prompt="(1) Please enter the tag ID: " )

 which_screen <- readline(prompt="(2) If you are using multiple screens, which screen do you wish to measure the data from? " )

 screen(as.numeric(as.character(which_screen)), FALSE) #Note we need to specify erase = FALSE.  Default is true.  If the FALSE statement is not included, the identify function will still work functionally, but will give erroneous values.  

 readline(prompt="(3) Please select the value of interest (press enter first) " )

 row_numbers <- identify(x$Date_Time, x$Power, n=1, plot=FALSE)

 values <- data.frame(matrix(1:3,1)) #Note: create a dataframe to store data; 3 columns, 1 row.

 names(values) <- c("Tag_id", "Power", "Date_Time") #Note: supply the names for the columns created above.

 values$Tag_id <- Tag

 values$Power <- x$Power[row_numbers]

 values$Date_Time <- x$Date_Time[row_numbers]

 return(values)

}

```

We can then apply the measuring function to the data.  You will be prompted to enter 1) the tag ID, 2) which screen (antenna) you want to extract the time from, and 3) press enter, then use your cursor to select the point on the plot that you want to extract the time stamp for.  Picking the last observation, for example, will give you the time of last detection.  If you want departure time, you could instead pick the point on the plot where the bird appears to become active and depart, which may be before the final detection.

```{r apply measure function, eval = FALSE}

departure_time <- NULL
departure_time <- rbind(departure_time, get_data(tag_155))
departure_time <- rbind(departure_time, get_data(tag_155)) # can do it again to get info from a separate screen
departure_time # print the selected time
#write.csv(departure_time, file = "DepartureTime_Tag155.csv", row.names = FALSE) # write to file, if you want
close.screen(all=T) # close all screens - will have to start over again.

```

